FOODiE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-10-09 papers/announcement/manuscript.tex

The other ODE solvers classification concerns with explicit or implicit nature of the schemes employed. Briefly, an explicit scheme computes the next step approximation using the previously computed steps at most to the current time, whereas an implicit scheme uses also the next step approximation (that is the unknown), thus it requires extra computations. The implicit approach is of practical use for \emph{stiff} systems where the usage of explicit schemes could require an extremely small time step to evolve in a \emph{stable} way the solution. Currently, FOODiE provides only explicit solvers.

@@

\begin{itemize}
  \item one-step schemes:
    \begin{itemize}
      \item explicit forward Euler scheme, it being $1^{st}$ order accurate;
      \item explicit Runge-Kutta schemes:
        \begin{itemize}
          \item TVD/SSP Runge-Kutta schemes:
            \begin{itemize}
              \item 2-stages, it being $2^{nd}$ order accurate;
              \item 3-stages, it being $3^{rd}$ order accurate;
              \item 5-stages, it being $4^{th}$ order accurate;
              \end{itemize}
          \item low storage Runge-Kutta schemes:
            \begin{itemize}
              \item 5-stages 2N registers schemes, it being $4^{th}$ order accurate;
              \end{itemize}
          \end{itemize}
      \end{itemize}
  \item multi-step schemes:
    \begin{itemize}
      \item Adams-Bashforth schemes:
        \begin{itemize}
          \item 2-steps, it being $2^{nd}$ order accurate;
          \item 3-steps, it being $3^{rd}$ order accurate;
          \item 4-steps, it being $4^{th}$ order accurate;
          \end{itemize}
        \item Leapfrog schemes\footnote{Leapfrog method is a 2-steps solver being, in general $2^{nd}$ order accurate, but mostly \emph{unstable}. It is well suited for periodic/oscillatory problems, thus the error is generally categorized accordingly to the phase and amplitude errors: the unfiltered scheme provides a $2^{nd}$ order error on phase approximation while the amplitude error is null. However, the unfiltered scheme is mostly unstable, thus many filters have been devised. In general, the filters retain the accuracy on the phase error, while affect the amplitude one.}:
        \begin{itemize}
          \item 2-steps unfiltered, it being $2^{nd}$ order accurate, but mostly \emph{unstable};
          \item 2-steps Robert-Asselin filtered, it being $1^{st}$ order accurate (on \emph{amplitude error});
          \item 2-steps Robert-Asselin-Williams filtered, it being $3^{rd}$ order accurate (on \emph{amplitude error});
          \end{itemize}
      \end{itemize}
  \end{itemize}

@@

Adams-Bashforth methods belong to the more general (linear) multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residuals function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

@@

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODiE main module},label={list:main_module}]
use foodie, only: integrand, &
                  adams_bashforth_integrator, &
                  euler_explicit_integrator, &
                  leapfrog_integrator, &
                  ls_runge_kutta_integrator, &
                  tvd_runge_kutta_integrator
! or simply
use foodie
\end{lstlisting}

@@

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{adams\_bashforth\_integrator} type},label={list:ab_solver}]
type :: adams_bashforth_integrator
  private
  integer(I_P)           :: steps=0 ! Number of time steps.
  real(R_P), allocatable :: b(:)    ! \(b\) coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    final                         :: finalize
endtype adams_bashforth_integrator
\end{lstlisting}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit Adams-Bashforth class},label={list:ab_integrate}]
subroutine integrate(self, U, previous, Dt, t)
class(adams_bashforth_integrator), intent(IN)    :: self         ! Actual AB integrator.
class(integrand),                  intent(INOUT) :: U            ! Field to be integrated.
class(integrand),                  intent(INOUT) :: previous(1:) ! Previous time steps solutions.
real(R_P),                         intent(IN)    :: Dt           ! Time steps.
real(R_P),                         intent(IN)    :: t(:)         ! Times.
integer(I_P)                                     :: s            ! Steps counter.
do s=1, self%steps
  U = U + previous(s)%t() * (Dt * self%b(s))
enddo
do s=1, self%steps - 1
  previous(s) = previous(s + 1)
enddo
previous(self%steps) = U
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third are the previous time steps solutions, the fourth is the time step used and the fifth is an array of the time values of the steps considered for the current one-step-over integration that are passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

It is worth noting that the method also performs the cyclic update of the previous time steps solutions memory registers.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of a Leapfrog integrator},label={list:use_lf_solver}]
real               :: times(1:2)
type(my_integrand) :: filter_displacement
type(my_integrand) :: my_field
type(my_integrand) :: previous(1:2)
call integrator%integrate(U=my_field, previous=previous, filter=filter_displacement, Dt=Dt, t=times)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT, \emph{previous} are the memory registers where previous time steps solutions are saved, \emph{filter\_displacement} is the register necessary for computing the eventual displacement of the applied filter and \emph{times} are the time at each 2 steps considered for the current one-step-over integration.

@@

Table \ref{tab:oscillation} summarizes the errors analysis.

@@

  \caption{Oscillation test errors analysis}\label{tab:oscillation}

@@

@@

>>>>

The other ODE solvers classification concerns with explicit or implicit nature of the schemes employed. Briefly, an explicit scheme computes the next step approximation using the previously computed steps at most to the current time, whereas an implicit scheme uses also the next step approximation (that is the unknown), thus it requires extra computations. The implicit approach is of practical use for \emph{stiff} systems where the usage of explicit schemes could require an extremely small time step to evolve in a \emph{stable} way the solution. Mixing together explicit and implicit schemes it is possible to build a family of \emph{predictor-corrector} methods: using an explicit scheme to predict a guess for the next step approximation it is possible to use an implicit method for correcting this guess. Currently, FOODiE provides explicit solvers and predictor-correct ones.

@@

\begin{itemize}
  \item one-step schemes:
    \begin{itemize}
      \item explicit forward Euler scheme, it being $1^{st}$ order accurate;
      \item explicit Runge-Kutta schemes:
        \begin{itemize}
          \item TVD/SSP Runge-Kutta schemes:
            \begin{itemize}
              \item 2-stages, it being $2^{nd}$ order accurate;
              \item 3-stages, it being $3^{rd}$ order accurate;
              \item 5-stages, it being $4^{th}$ order accurate;
              \end{itemize}
          \item low storage Runge-Kutta schemes:
            \begin{itemize}
              \item 5-stages 2N registers schemes, it being $4^{th}$ order accurate;
              \end{itemize}
          \end{itemize}
      \end{itemize}
  \item multi-step schemes:
    \begin{itemize}
      \item explicit Adams-Bashforth schemes:
        \begin{itemize}
          \item 2-steps, it being $2^{nd}$ order accurate;
          \item 3-steps, it being $3^{rd}$ order accurate;
          \item 4-steps, it being $4^{th}$ order accurate;
          \end{itemize}
      \item implicit Adams-Moulton schemes:
        \begin{itemize}
          \item 1-step, it being $2^{nd}$ order accurate;
          \item 2-steps, it being $3^{rd}$ order accurate;
          \item 3-steps, it being $4^{th}$ order accurate;
          \end{itemize}
      \item predictor-corrector Adams-Bashforth-Moulton schemes:
        \begin{itemize}
          \item 1-step, it being $2^{nd}$ order accurate;
          \item 2-steps, it being $3^{rd}$ order accurate;
          \item 3-steps, it being $4^{th}$ order accurate;
          \end{itemize}
      \item explicit Leapfrog schemes\footnote{Leapfrog method is a 2-steps solver being, in general $2^{nd}$ order accurate, but mostly \emph{unstable}. It is well suited for periodic/oscillatory problems, thus the error is generally categorized accordingly to the phase and amplitude errors: the unfiltered scheme provides a $2^{nd}$ order error on phase approximation while the amplitude error is null. However, the unfiltered scheme is mostly unstable, thus many filters have been devised. In general, the filters retain the accuracy on the phase error, while affect the amplitude one.}:
        \begin{itemize}
          \item 2-steps unfiltered, it being $2^{nd}$ order accurate, but mostly \emph{unstable};
          \item 2-steps Robert-Asselin filtered, it being $1^{st}$ order accurate (on \emph{amplitude error});
          \item 2-steps Robert-Asselin-Williams filtered, it being $3^{rd}$ order accurate (on \emph{amplitude error});
          \end{itemize}
      \end{itemize}
  \end{itemize}

@@

Adams-Bashforth methods belong to the more general (linear) explicit multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residuals function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

@@

\subsection{The implicit Adams-Moulton class of schemes}

Adams-Moulton methods belong to the more general (linear) implicit multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residuals function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

In general, the Adams-Moulton schemes provided by FOODiE library are written by means of the following algorithm (for only implicit schemes):

\begin{equation}
  U\left(t^{n+N_s}\right) = U\left(t^{n+N_s-1}\right) +\Delta t \sum_{s=0}^{n+N_s-1}{ b_s \cdot R\left[t^{n+s}, U\left(t^{n+s}\right)\right]} + b_{N_s}\cdot R\left[t^{n+N_s}, U\left(t^{n+N_s}\right)\right]
\label{eq:AM}
\end{equation}
where $N_s$ is the number of time steps considered and $b_s$ are the linear coefficients selected.

Currently FOODiE provides 1, 2, and 3 steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s$ coefficients are reported in table \ref{tab:AM-coeff}.

\begin{table}[!ht]
  \centering
  \caption{Implicit Adams-Moulton coefficients\label{tab:AM-coeff}}
  \resizebox{0.30\textwidth}{!}{%
  \begin{tabular}{c|cccc}
    $N_s$ & $b_0$           & $b_1$           & $b_2$           & $b_3$          \\
    \hline
    1     & $\frac{1}{2}$   & $\frac{1}{2}$   &       /         &       /        \\
    2     & $-\frac{1}{12}$ & $\frac{8}{12}$  & $\frac{5}{12}$  &       /        \\
    3     & $\frac{1}{24}$  & $-\frac{5}{24}$ & $\frac{19}{24}$ & $\frac{9}{24}$ \\
  \end{tabular}}
\end{table}

Similarly to the Runge-Kutta and Adams-Bashforth classes, the Adams-Moulton class also provides a fail-safe zero-step solver reverting back to the implicit backward Euler solver, that is useful for ROR-like frameworks.

It is worth noting that for $N_s>1$ the Adams-Moulton class of solvers is not \emph{self-starting}: the values of $U\left(t^{1}\right)$, $U\left(t^{2}\right)$, \dots, $U\left(t^{N_s-1}\right)$ must be provided. To this aim, a lower order multi-step scheme or an equivalent order one-step multi-stage scheme can be used.

\subsection{The predictor-corrector Adams-Bashforth-Moulton class of schemes}

Adams-Bashforth-Moulton methods belong to the more general (linear) predictor-corrector multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residuals function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

In general, the Adams-Bashforth-Moulton schemes provided by FOODiE library are written by means of the following algorithm:

\begin{equation}
  \begin{matrix}
    U\left(t^{n+N_s^p}\right)_p = U\left(t^{n+N_s^p-1}\right) +\Delta t \sum_{s=1}^{n+N_s^p}{ b_s^p \cdot R\left[t^{n+s-1}, U\left(t^{n+s-1}\right)\right]}   \\
    U\left(t^{n+N_s^c}\right)_c = U\left(t^{n+N_s^c-1}\right) +\Delta t \sum_{s=0}^{n+N_s^c-1}{ b_s^c \cdot R\left[t^{n+s}, U\left(t^{n+s}\right)\right]} + b_{N_s^c}^c\cdot R\left[t^{n+N_s^p}, U\left(t^{n+N_s^p}\right)_p\right]
  \end{matrix}
\label{eq:ABM}
\end{equation}
where $N_s^{p,c}$ is the number of time steps considered for the Adams-Bashforth predictor/Adams-Moulton corrector (respectively) and $b_s^{p,c}$ are the corresponding linear coefficients selected. Essentially, the Adams-Bashforth prediction $U\left(t^{n+N_s^p}\right)_p$ is corrected by means of the Adams-Moulton correction resulting in $U\left(t^{n+N_s^c}\right)_c$. In order to preserve the formal order of accuracy the relation $N_s^p=N_s^c+1$ always holds.

Currently FOODiE provides $N_s^c=1, 2, 3 \rightarrow N_s^c=2, 3, 4$ steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s^{p,c}$ coefficients are those reported in tables \ref{tab:AB-coeff} and \ref{tab:AM-coeff}.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODiE main module},label={list:main_module}]
use foodie, only: integrand, &
                  adams_bashforth_integrator, &
                  adams_moulton_integrator, &
                  adams_bashforth_moulton_integrator, &
                  euler_explicit_integrator, &
                  leapfrog_integrator, &
                  ls_runge_kutta_integrator, &
                  tvd_runge_kutta_integrator
! or simply
use foodie
\end{lstlisting}

@@

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time;
  \item \emph{update\_previous}: auto update (cyclically) previous time steps solutions.
  \end{itemize}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{adams\_bashforth\_integrator} type},label={list:ab_solver}]
type :: adams_bashforth_integrator
  private
  integer(I_P)           :: steps=0 ! Number of time steps.
  real(R_P), allocatable :: b(:)    ! \(b\) coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    procedure, pass(self), public :: update_previous
    final                         :: finalize
endtype adams_bashforth_integrator
\end{lstlisting}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit Adams-Bashforth class},label={list:ab_integrate}]
subroutine integrate(self, U, previous, Dt, t, autoupdate)
class(adams_bashforth_integrator), intent(IN)    :: self         ! Actual AB integrator.
class(integrand),                  intent(INOUT) :: U            ! Field to be integrated.
class(integrand),                  intent(INOUT) :: previous(1:) ! Previous time steps solutions.
real(R_P),                         intent(IN)    :: Dt           ! Time steps.
real(R_P),                         intent(IN)    :: t(:)         ! Times.
logical, optional,                 intent(IN)    :: autoupdate   ! Autoupdate previous time steps.
logical                                          :: autoupdate_  ! autoupdate previous time steps, dummy var.
integer(I_P)                                     :: s            ! Steps counter.
autoupdate_ = .true. ; if (present(autoupdate)) autoupdate_ = autoupdate
do s=1, self%steps
  U = U + previous(s)%t(t=t(s)) * (Dt * self%b(s))
enddo
if (autoupdate_) call self%update_previous(U=U, previous=previous)
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third are the previous time steps solutions, the fourth is the time step used, the fifth is an array of the time values of the steps considered for the current one-step-over integration that are passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time and the sixth is a logical flag for enabling/disabling the cyclic update of previous time steps solutions. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

It is worth noting that the method also performs the cyclic update of the previous time steps solutions memory registers. This can be disable passing \emph{autoupdate=.false.}: it is useful in the framework of predictor-corrector solvers.

\subsection{The implicit Adams-Moulton class of solvers}\label{subsec:solver_am}

The explicit Adams-Moulton class of solvers is exposed as a single derived type named \emph{adams\_moulton\_integrator}. This type provides three methods:

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time;
  \item \emph{update\_previous}: auto update (cyclically) previous time steps solutions.
  \end{itemize}

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_am_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an implicit Adams-Moulton integrator},label={list:define_am_solver}]
use FOODiE, only: adams_moulton_integrator
type(adams_moulton_integrator) :: integrator
\end{lstlisting}

Once an integrator of this type has been instantiated, it must be initialized before used, for example see listing \ref{list:init_am_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of initialization of an implicit Adams-Moulton integrator},label={list:init_am_solver}]
call integrator%init(steps=3)
\end{lstlisting}

In the listing \ref{list:init_am_solver} a 3-steps solver has been initialized. As a matter of facts, from the equation \ref{eq:AM} a solver belonging to this class is completely defined once the number of time steps adopted has been chosen. The complete definition of the \emph{adams\_moulton\_integrator} type is reported into listing \ref{list:am_solver}. As shown, the linear coefficients are stored as allocatable arrays the values of which are initialized by the \emph{init} method.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{adams\_moulton\_integrator} type},label={list:am_solver}]
type :: adams_moulton_integrator
  private
  integer(I_P)           :: steps=-1 ! Number of time steps.
  real(R_P), allocatable :: b(:)     ! \(b\) coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    procedure, pass(self), public :: update_previous
    final                         :: finalize
endtype adams_moulton_integrator
\end{lstlisting}

After the solver has been initialized it can be used for integrating an integrand field, as shown in listing \ref{list:use_am_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of an Adams-Moulton integrator},label={list:use_am_solver}]
real               :: times(1:3)
type(my_integrand) :: my_field
type(my_integrand) :: previous(1:3)
call integrator%integrate(U=my_field, previous=previous, Dt=Dt, t=times)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT, \emph{times} are the time at each 4 steps considered for the current one-step-over integration and \emph{previous} are the memory registers where previous time steps solutions are saved.

The complete implementation of the \emph{integrate} method of the implicit Adams-Moulton class of solvers is reported in listing \ref{list:am_integrate}.

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit Adams-Moulton class},label={list:am_integrate}]
subroutine integrate(self, U, previous, Dt, t, autoupdate)
class(adams_bashforth_integrator), intent(IN)    :: self         ! Actual AB integrator.
class(integrand),                  intent(INOUT) :: U            ! Field to be integrated.
class(integrand),                  intent(INOUT) :: previous(1:) ! Previous time steps solutions.
real(R_P),                         intent(IN)    :: Dt           ! Time steps.
real(R_P),                         intent(IN)    :: t(:)         ! Times.
logical, optional,                 intent(IN)    :: autoupdate   ! Autoupdate previous time steps.
logical                                          :: autoupdate_  ! autoupdate previous time steps, dummy var.
integer(I_P)                                     :: s            ! Steps counter.
autoupdate_ = .true. ; if (present(autoupdate)) autoupdate_ = autoupdate
if (autoupdate_) call self%update_previous(U=U, previous=previous)
if (self%steps>0) then
  U = previous(self%steps) + U%t(t=t(self%steps) + Dt) * (Dt * self%b(self%steps))
  do s=0, self%steps - 1
    U = U + previous(s+1)%t(t=t(s+1)) * (Dt * self%b(s))
  enddo
  if (autoupdate_) call self%update_previous(U=U, previous=previous)
else
  U = U + U%t(t=t(s+1)) * (Dt * self%b(0))
endif
return
endsubroutine integrate
\end{lstlisting}

This method takes six arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third are the previous time steps solutions, the fourth is the time step used, the fifth is an array of the time values of the steps considered for the current one-step-over integration that are passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time and the sixth is a logical flag for enabling/disabling the cyclic update of previous time steps solutions. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

It is worth noting that the method also performs the cyclic update of the previous time steps solutions memory registers. This can be disable passing \emph{autoupdate=.false.}: it is useful in the framework of predictor-corrector solvers.

\subsection{The predictor-corrector Adams-Bashforth-Moulton class of solvers}\label{subsec:solver_abm}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of a Leapfrog integrator},label={list:use_lf_solver}]
real               :: times(1:2)
type(my_integrand) :: filter_displacement
type(my_integrand) :: my_field
type(my_integrand) :: previous(1:2)
call integrator%integrate(U=my_field, previous=previous, filter=filter_displacement, Dt=Dt, &
                          t=times)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT, \emph{previous} are the memory registers where previous time steps solutions are saved, \emph{filter\_displacement} is the register necessary for computing the eventual displacement of the applied filter and \emph{times} are the time at each 2 steps considered for the current one-step-over integration.

@@

Tables \ref{tab:oscillation_errors_explicit} and \ref{tab:oscillation_errors_pred_corr} summarize the errors analysis.

@@

  \caption{Oscillation test: errors analysis of explicit solvers}\label{tab:oscillation_errors_explicit}

@@

\begin{table}[!ht]
  \centering
  \caption{Oscillation test: errors analysis of predictor-corrector solvers}\label{tab:oscillation_errors_pred_corr}
  \begin{subtable}[b]{0.45\textwidth}
    \centering
    \caption{Adams-Bashforth-Moulton 1 step}\label{tab:oscillation-abm-1}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      \toprule
      {\sc Time Step} & {\sc Error X} & {\sc Error Y} & {\sc Order X} & {\sc Order Y} \\
      \hline
      5000.0          &  0.241E+20    &  0.266E+20    & /             & /             \\
      2500.0          &  0.664E+11    &  0.716E+11    & 28.44         & 28.47         \\
      1250.0          &  0.952E+06    &  0.100E+07    & 16.09         & 16.12         \\
       625.0          &  0.413E+04    &  0.407E+04    &  7.85         &  7.95         \\
       320.0          &  0.387E+03    &  0.383E+03    &  3.54         &  3.53         \\
       100.0          &  0.145E+03    &  0.145E+03    &  0.84         &  0.83         \\
      \bottomrule
    \end{tabular}}
  \end{subtable}\quad%
  \begin{subtable}[b]{0.45\textwidth}
    \centering
    \caption{Adams-Bashforth-Moulton 2 steps}\label{tab:oscillation-abm-2}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      \toprule
      {\sc Time Step} & {\sc Error X} & {\sc Error Y} & {\sc Order X} & {\sc Order Y} \\
      \hline
      5000.0          &  0.704E+01    &  0.701E+01    & /             & /             \\
      2500.0          &  0.392E+01    &  0.395E+01    & 0.84          & 0.83          \\
      1250.0          &  0.148E+01    &  0.150E+01    & 1.40          & 1.39          \\
       625.0          &  0.526E+00    &  0.534E+00    & 1.49          & 1.49          \\
       320.0          &  0.193E+00    &  0.196E+00    & 1.50          & 1.50          \\
       100.0          &  0.338E-01    &  0.342E-01    & 1.50          & 1.50          \\
      \bottomrule
    \end{tabular}}
  \end{subtable}\\
  \begin{subtable}[b]{0.45\textwidth}
    \centering
    \caption{Adams-Bashforth-Moulton 3 steps}\label{tab:oscillation-abm-3}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      \toprule
      {\sc Time Step} & {\sc Error X} & {\sc Error Y} & {\sc Order X} & {\sc Order Y} \\
      \hline
      5000.0          &  0.457E+01    &  0.464E+01    & /             & /             \\
      2500.0          &  0.656E+00    &  0.654E+00    & 2.80          & 2.83          \\
      1250.0          &  0.100E+00    &  0.987E-01    & 2.71          & 2.73          \\
       625.0          &  0.169E-01    &  0.167E-01    & 2.56          & 2.56          \\
       320.0          &  0.314E-02    &  0.310E-02    & 2.52          & 2.51          \\
       100.0          &  0.171E-03    &  0.169E-03    & 2.50          & 2.50          \\
      \bottomrule
    \end{tabular}}
  \end{subtable}\quad%
  \begin{subtable}[b]{0.45\textwidth}
    \centering
    \caption{Adams-Bashforth-Moulton 4 steps}\label{tab:oscillation-abm-4}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      \toprule
      {\sc Time Step} & {\sc Error X} & {\sc Error Y} & {\sc Order X} & {\sc Order Y} \\
      \hline
      5000.0          &  0.229E+01    &  0.225E+01    & /             & /             \\
      2500.0          &  0.119E+00    &  0.118E+00    & 4.26          & 4.25          \\
      1250.0          &  0.825E-02    &  0.833E-02    & 3.85          & 3.83          \\
       625.0          &  0.671E-03    &  0.681E-03    & 3.62          & 3.61          \\
       320.0          &  0.631E-04    &  0.640E-04    & 3.53          & 3.53          \\
       100.0          &  0.107E-05    &  0.108E-05    & 3.51          & 3.51          \\
      \bottomrule
    \end{tabular}}
  \end{subtable}\\
\end{table}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-adams-bashforth-moulton-1.png}
    \caption{1 step solution}\label{fig:results-oscillation-adams-bashforth-moulton-1}
  \end{subfigure}\quad%
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-adams-bashforth-moulton-2.png}
    \caption{2 steps solution}\label{fig:results-oscillation-adams-bashforth-moulton-2}
  \end{subfigure}\\
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-adams-bashforth-moulton-3.png}
    \caption{3 steps solution}\label{fig:results-oscillation-adams-bashforth-moulton-3}
  \end{subfigure}\quad%
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-adams-bashforth-moulton-4.png}
    \caption{4 steps solution}\label{fig:results-oscillation-adams-bashforth-moulton-4}
  \end{subfigure}
  \caption{Oscillation equations solutions computed by means of Adams-Bashforth-Moulton solvers}\label{fig:results-oscillation-adams-bashforth-moulton}
\end{figure}
\clearpage
